The noise-marginalized optimal statistic: A robust hybrid frequentist-Bayesian statistic for a stochastic gravitational-wave background in pulsar timing arrays

### 2018-05-07 manuscript/manuscript.tex
edits to abstract and introduction

\begin{abstract}
Observations have revealed that nearly all galaxies 
contain supermassive black holes (SMBHs) at their centers. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these GW sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator for the amplitude of a GWB 
that specifically looks for the spatial correlations induced by the GWB in PTA data. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of a GWB, 
and we use the noise-marginalized optimal statistic to assess the 
significance of the spatial correlations 
by comparing monopole, dipole, and Hellings-Downs spatial correlations 
and performing sky scrambles.
\end{abstract}

@@

PTAs primarily use Bayesian data analysis to compare the inferred probabilities 
of various models for the pulsar timing residuals, 
including one where the residuals contain evidence of the GWB 
\citep{vlm+2009,lah+2013}. 
%Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
Bayesian inference is a powerful tool because 
it performs a non-linear fit and 
accounts for degeneracies between parameters. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of Hellings-Downs spatial correlations --
the ``smoking gun'' of the GWB.

The significance of the GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude \citep{abc+2009,demorest+2013,ccs+2015}. 
%It is an important complement to Bayesian detection techniques 
%because it provides an independent detection procedure, 
%and it is significantly faster to compute. 
%In particular, different spatial correlations can be tested with the optimal statistic 
%in less than a second, 
%whereas a Bayesian analysis with spatial correlations requires 
%several weeks to run on a supercomputing cluster.
Not only does it provide an independent detection procedure, complementing a more robust Bayesian analysis, but its calculation requires significantly less time to compute. In particular, the optimal statistic produces results for various spatial correlation functions, a bottleneck within the more traditional stochastic GWB analysis, within seconds; a full Bayesian analysis including correlations has to run for several weeks on a supercomputing cluster.

However, the optimal statistic gives biased results 
when the pulsars have significant red noise 
due to the strong covariance between the individual red noise parameters 
and the GWB amplitude. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes more than the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic with any choice of overlap-reduction function (ORF), 
allowing for quick testing of various spatial correlations. 
This technique is used to perform the frequentist searches 
for common red signals with monopole, dipole, and Hellings-Downs spatial correlations 
in the NANOGrav 11-year data set \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed using fixed noise. 
In Sec.~\ref{sec:spatial} we determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and Hellings-Downs spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we determine how well the noise-marginalized optimal statistic 
can assess the significance of a GW detection by performing sky scrambles \citep{cs2016, tlb+2017}. 
%We also compare our sky scramble results to the Bayesian analysis of sky scrambles done by \citet{tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
and discuss future applications of the noise-marginalized optimal statistic.

>>>>

\begin{abstract}
Observations have revealed that nearly all galaxies 
contain supermassive black holes (SMBHs) at their centers. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator of the amplitude of the GWB 
that specifically looks for the spatial correlations induced by the GWB in PTA data. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of the GWB, 
and we use the noise-marginalized optimal statistic 
to compare the significance of monopole, dipole, and Hellings-Downs spatial correlations 
and perform sky scrambles.
\end{abstract}

@@

PTAs primarily use Bayesian data analysis to compare the inferred probabilities 
of various models for the residuals, 
including one where they contain evidence of the GWB 
\citep{vlm+2009,lah+2013}. 
%Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
Bayesian inference is a powerful tool because 
it performs a non-linear fit and 
accounts for degeneracies between parameters. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of Hellings-Downs spatial correlations --
the ``smoking gun'' of the GWB.

The significance of the GWB can also be assessed using the 
optimal statistic, a frequentist estimator for the GWB amplitude \citep{abc+2009,demorest+2013,ccs+2015}. 
%It is an important complement to Bayesian detection techniques 
%because it provides an independent detection procedure, 
%and it is significantly faster to compute. 
%In particular, different spatial correlations can be tested with the optimal statistic 
%in less than a second, 
%whereas a Bayesian analysis with spatial correlations requires 
%several weeks to run on a supercomputing cluster.
Not only does it provide an independent detection procedure, complementing a more robust Bayesian analysis, but it requires significantly less time to compute. In particular, the optimal statistic produces results for various spatial correlation functions within seconds; a full Bayesian analysis including correlations has to run for many weeks on a supercomputing cluster.

However, the optimal statistic gives biased results 
when the pulsars have significant red noise 
due to the strong covariance between the individual red noise parameters 
and the GWB amplitude. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes more than the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic for any choice of spatial correlations 
simply by changing the overlap-reduction function (ORF). 
This technique is used to perform the frequentist searches 
for common red signals with monopole, dipole, and Hellings-Downs spatial correlations 
in the NANOGrav 11-year data set \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed with fixed noise. 
In Sec.~\ref{sec:spatial} we determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and Hellings-Downs spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scrambles, 
which assess the significance of Hellings-Downs spatial correlations 
by scrambling the pulsars' sky positions \citep{cs2016, tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
and discuss future applications of the noise-marginalized optimal statistic.
